Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-26 paper/paper.tex
DFA first half

A key construct in formal language theory is the Brzozowski derivative
\cite{brzozowski}. Given a string $w$ the (left) derivative of a
language $L$ with respect to $w$, written $\partial_w L$ is the language
\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
This semantics is given using the linear implication $w \tol
L$. Expressing Brzozowski's calculation of the derivative of a regular
expression relies on additional axioms to the type theory.\max{axioms
  to carry out Brzozowski's derivative calculations plz.}

\begin{align*}
  c\lto c &\equiv I\\
  c\lto d &\equiv 0\\
  c\lto I &\equiv 0\\
  c\lto 0 &\equiv 0\\
  c\lto (A \otimes B) &\equiv (c\lto A) \otimes B + \nu(A) \otimes (c\lto B)\\
  c\lto A^* &\equiv (c\lto A) \otimes A^*\\
  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
\end{align*}

\pedro{The type family $\nu$ above is equal to $I$ if its argument
  parses the empty string and zero otherwise. In order to define
  it we should probably add an extra axiom, right?}
\steven{$\nu$ can be derived using $\amp$, $\nu(A) = I \amp A$}

\steven{In addition to Pedro's axioms:

$c, c'$ characters. $g$ any grammar.

\begin{gather}
  I \lto c \cong c \\
  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
\end{gather}

For axiom $(1)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity. We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$. For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.

Axiom $(2)$ might be admissible, but I'm unsure.

I think for Kleene star we need $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$. This should derivable utilizing the product axiom and $g^{*} \cong I \oplus (g \otimes g^{*})$, or is that sort of isomorphism opaque to us right now.

To make the DFA proof go through I think we additionally need to axiomatically make the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso. Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,

but this last bit isn't just Brzozowski's definitions

}
Using the $G$ modality, we can internalize some of the external
reasoning about grammars into the language of non-linear types. Recall
that the denotation of $G A$ is the set of parses of the empty
string. As is well known in formal language theory, all the parsing
behavior of a grammar can be reduced to reasoning about parsing of the
empty string by using the derivative of a grammar. For instance, the
set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
and so the type of all parses of a grammar can be defined as
$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
\tol A)$ are equal.\max{I think we should add a propositional
  truncation to the type theory.}  Similarly, the type of \emph{total
parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
function that selects a parse for every possible string. A
\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
where the parser is allowed to fail for any reason. This has the
downside that the always failing semi-parser $\lambda _. \textrm{in}_r
()$ is valid for any grammar. If we were working with a full
\emph{bunched} type theory, we would be able to express the more
precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
implication.

@@

An NFA has finitely many states, so we may number them $Q = \{q_{1}, \dots, q_{N}\}$. For each $q_{j}$ we introduce
a mutually recursive term $Y_{j}$ to represent the states of the NFA within this grammar presentation. We represent the transition function $\delta$ in the right hand side of each $Y_{j}$ definition. For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,

@@

\steven{Somewhere mention that $\epsilon$ corresponds to $I$, characters $c \in \Sigma$ correspond to grammars $c$, etc}

and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
denotes an $\epsilon$-transition. A
disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

@@

This construction

@@

\[
  \inferrule{
    \Delta \vdash e : NFA(g \oplus g') \\
    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{~where } j \not \in \{s, e\} \\
    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{~where } j \not \in \{s, e\} \\
    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(e) : g \oplus g'}
\]

@@

\[
  \inferrule
  {
    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{s} \\
    \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = X_{j}' \vdash x_{j} : X_{j}', j \not \in \{s, e\} \\
    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash x_{e} : I = X_{e}'
  }
  {\Gamma ; \Delta \vdash \mathsf{mfold}(u) : NFA(g \oplus g')}
\]
where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$. \steven{Need a better way to display these defs without blowing up the $\mu$ terms} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$. Similarly define an embedding $\beta$ for an embedding of $NFA(g')$. We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with maps $\alpha$ and $\beta$, respectively.

@@

Recall that a DFA is given as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$ in the same manner as an NFA.\ When encoding a DFA as a mutually recursive grammar, we use nearly the same process as when we encoded an NFA, with one subtle change. When we encoded NFAs, we were able to ensure we had a single accepting state by connecting all of the accepting states $q_{j} \in F$ to a single accepting state $Y_{e}$ with $\epsilon$-transitions. However, $\epsilon$-transitions are not permitted in a DFA, as they introduce a source of nondeterminism. Therefore we need a scheme to encode which states are accepting. The change we need to make is pretty lightweight, we add the grammar $I$ as a summand to distinguish which states are accepting.

\[
  DFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle \hat{Y} \rangle \\
        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
        \end{pmatrix}.Y_{s}
\]
where
\[
\hat{Y} ::= \hat{Y_{0}} ~|~ I \oplus \hat{Y_{0}}
\]
\[
  \hat{Y_{0}} ::= c \otimes Y ~|~ Y ~|~ \hat{Y_{0}} \oplus \hat{Y_{0}}, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
\]

@@

some fluff to lead into the thoerem

\begin{theorem}
  For every regular grammar $g$, there is a DFA $D$ that is weakly equivalent to $g$.
\end{theorem}

\begin{proof}
  \steven{This is currently word salad}

  DFA construction

  $DFA(g) \vdash g$ : just fold over the defs in $DFA(g)$ and have every $Y_{w}$ eliminate to $w \lto g$. This is doable bc when eliminating $Y_{w'}$ you will hold terms of type $a \otimes (w \lto g)$. These $w \lto g$ terms might not be exactly what you want, but they are weakly equivalent to some derivative that you do want! How do you know that they're weakly equivalent? if $Y_{w'}$ transitions to $Y_{w}$ via character $a$, then we know that $a \lto (w \lto g)$ is weakly equivalent to $a \lto (w' \lto g)$ \emph{by construction of the DFA}. This argument entirely hinges on the method by which we defined the DFA

  $g \vdash DFA(g)$
  Idea is to us $\top \cong \text{Char}^{*}$ and $A^{*} \cong \Sigma(n) A^{\otimes n}$ and $A \cong \top \amp A$ to turn a parse into a pair of a parse and a finite string. And then this finite string can be walked through the DFA to read off the parse, and you can show its accepting. Details currently a little shaky, and encapsulated in a bad picture of a whiteboard. \steven{Gonna has this out with Max in meeting}
\end{proof}

@@

  \steven{There a weird spacing issues between these align envs. Maybe wrap these in a figure env so that spacing is better}

\begin{align*}
  r_{\oplus}(s_{\oplus}(e))
  & =_{\mu \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus \mu Z_{s}' \\
  &
    \begin{aligned}
      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{ & \\
    \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, & \\
    \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
    \})))
    \end{aligned} \\
  &
    \begin{aligned}
    =_{\oplus \eta}
    \textbf{case}~e'~\{ & \\
    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
    \}))
    \end{aligned} \\
  &
    \begin{aligned}
    =
    \textbf{case}~e'~\{ & \\
    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
    \}
    \end{aligned} \\
  &
    \begin{aligned}
    =_{\mu \beta}
    \textbf{case}~e'~\{ & \\
    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\textbf{in}_{g} \phi^{-1}(\mathsf{cons}_{\mu(X_{i}=\hat{X}_{i}).X_{s}}u')), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v')) \\
      \} \\
      & \text{~for }u' : \mu X_{s}', v' : \mu Z_{s}'
    \end{aligned} \\
  &
    \begin{aligned}
    =_{\oplus \beta}
    \textbf{case}~e'~\{ & \\
    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha \phi \phi^{-1}(\mathsf{cons}(u')), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v')) \\
      \}
    \end{aligned} \\
  &
    \begin{aligned}
    =
    \textbf{case}~e'~\{ & \\
    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha (\mathsf{cons}(u')), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v')) \\
      \}
    \end{aligned} \\
   &
    \begin{aligned}
    =_{\mu \beta}
    \textbf{case}~e'~\{ & \\
    & \textbf{in}_{\mu X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{\mu X_{s}'} u) , \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v) \\
      \}
    \end{aligned} \\
  & =_{\oplus \eta} \mathsf{cons}(e') \\
  & =_{\mu \beta} e
\end{align*}

@@

\[
  NFA(g \otimes g') = \mu
  \begin{pmatrix}
    Y_{s} = X_{s}' \\
    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
    X_{e}' = Z_{s}' \\
    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
    Z_{e}' = Y_{e} \\
    Y_{e} = I
  \end{pmatrix}
\]

>>>>

A key construct in formal language theory is the Brzozowski derivative
\cite{brzozowski}. Given a string $w$ the (left) derivative of a
language $L$ with respect to $w$, written $\partial_w L$ is the language
\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
This semantics is given using the linear implication $w \tol
L$. Expressing Brzozowski's calculation of the derivative of a regular
expression relies on additional axioms to the type theory, presented in \cref{fig:brzozowskideriv}.

\begin{figure}
\label{fig:brzozowskideriv}
\begin{align*}
  c\lto c &\equiv I\\
  c\lto d &\equiv 0\\
  c\lto I &\equiv 0\\
  c\lto 0 &\equiv 0\\
  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
\end{align*}
\caption{Equality for Brzozowski Derivatives}
\end{figure}

Using the $G$ modality, we can internalize some of the external
reasoning about grammars into the language of non-linear types. Recall
that the denotation of $G A$ is the set of parses of the empty
string. As is well known in formal language theory, all the parsing
behavior of a grammar can be reduced to reasoning about parsing of the
empty string by using the derivative of a grammar. For instance, the
set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
and so the type of all parses of a grammar can be defined as
$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
\tol A)$ are equal.\max{I think we should add a propositional
  truncation to the type theory.}  Similarly, the type of \emph{total
parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
function that selects a parse for every possible string. A
\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
where the parser is allowed to fail for any reason. This has the
downside that the always failing semi-parser $\lambda _. \textrm{in}_r
()$ is valid for any grammar. If we were working with a full
\emph{bunched} type theory, we would be able to express the more
precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
implication.

@@

An NFA has finitely many states, so we may number them $Q = \{q_{1}, \dots, q_{N}\}$. For each $q_{j}$ we introduce
a mutually recursive term $Y_{j}$ to represent the states of the NFA within this grammar presentation. We represent the transition function $\delta$ in the right hand side of each $Y_{j}$ definition. For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinSigTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,

@@

and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
denotes an $\epsilon$-transition. Note that here each $c$ represents a grammar that is inhabited with only a trivial parse of the single character $c$; likwise, and $\epsilon$-transtion corresponds to a transition with label $I$ and $I \otimes Y = Y$ because $I$ is the unit for the operation $\otimes$. A
disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

@@

@@

\[
  \inferrule{
    \Delta \vdash e : NFA(g \oplus g') \\
    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{~where } j \not \in \{s, e\} \\
    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{~where } j \not \in \{s, e\} \\
    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(y_{s}.p_{s},\dots,y_{e}.p_{e})(e) : g \oplus g'}
\]

@@

\[
  \inferrule
  {
    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{s} \\
    \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{j}' \vdash x_{j} : \mu(X_{j} = \hat{X}_{j}, \forall j).X_{j}', j \not \in \{s, e\} \\
    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash \mathsf{cons}(x_{e}) : I = \mu(\dots).Y_{e} = \mu(\dots).X_{e}'
  }
  {\Gamma ; \Delta \vdash \mathsf{mfold}(x_{s}.\mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s}), x_{1}.x_{1}, \dots, x_{e}.x_{e}))(u) : NFA(g \oplus g')}
\]
where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$. \steven{Need a better way to display these defs without blowing up the $\mu$ terms} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$. Similarly define an embedding $\beta$ for an embedding of $NFA(g')$. We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with maps $\alpha$ and $\beta$, respectively.

@@

Recall that a DFA is given as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$ in the same manner as an NFA.\ When encoding a DFA as a mutually recursive grammar, we use nearly the same process as when we encoded an NFA, with one subtle change. When we encoded NFAs, we were able to ensure we had a single accepting state by connecting all of the accepting states $q_{j} \in F$ to a single accepting state $Y_{e}$ with $\epsilon$-transitions. However, $\epsilon$-transitions are not permitted in a DFA, as they introduce a source of nondeterminism. Therefore we need a scheme to encode which states are accepting. The changes we need to make is pretty lightweight. First, we add the grammar $I$ as a summand to distinguish which states are accepting. Second, each state must have a transition for every character in the alphabet.

\[
  DFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle \hat{Y} \rangle \\
        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
        \end{pmatrix}.Y_{s}
\]
where
\[
\hat{Y} ::= \hat{Y_{0}} ~|~ I \oplus \hat{Y_{0}}
\]
\[
  \hat{Y_{0}} ::= \bigoplus_{c : \Sigma} c \otimes Y, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
\]

@@

In the definition of a DFA, we make use of a sum ranging over all characters in the alphabet. Viewed as a machine, this is intened to have the effect of being able to run any arbitrary string input through the automaton to reach some well-defined, but not necessarily accepting, state. In our type theory, this would amount to the following entailment being derivable,

\[
p : \top \vdash \mathsf{run}(p) : \mu \begin{pmatrix}
        Y_{s} = \langle \hat{Y} \rangle \\
        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
        \end{pmatrix}.\bigoplus_{k} Y_{k}
\]

for some term $\mathsf{run}(p)$ that might depend on the structure of the DFA in question. That is, any input should be runnable through the DFA to return \emph{some} state. This entailment holds in the intended model, but a priori this does not hold in the type theory. Noting this point motivates taking $\top \cong (\bigoplus_{c} c)^{*}$ as an axiom. This syntactically reflects the semantic notion that parses correspond to finite strings over a fixed alphabet. Such an axiom is especially justified when crafting syntax to reason about parses of finite input strings; further, we conjecture that without this axiom we may find models of the type theory that correspond to parses of infinite sequences of characters. In this paper, we shall focus on the expressivity of the syntax to reason about parses of strings, and thus we assume $\top \cong (\bigoplus_{c} c)^{*}$.

Additionally, we assume that the derivation from $c \otimes (c \lto A)$ to $(c \otimes \top) \amp A$ is an isomorphism. Specifically, the term $\phi$ defined below is assumed to have an inverse,

Axiomatically adding an inverse for $\phi$ doesn't greatly affect our semantics. In fact, it describes a property that is simply true in the intended string model: ``if a parse of character $c$ and then a parse of any word completes to a parse of $A$, then that word was actually in fact of type $c \lto A$''. This axiom, the one desribed in the paragraph above, and the ones given in \cref{fig:brzozowskideriv} are not meant to anything particularly deep. They are a means of reflecting the intended semantic properties in our syntax.

Utilizing the reflected properties we can derive the following isomorphism,

\begin{align*}
  A
  & \cong \top \amp A \\
  & \cong (\bigoplus_{c} c)^{*} \amp A \\
  & \cong (I \oplus \bigoplus_{c} (c \otimes \top)) \amp A \\
  & \cong (I \amp A) \oplus \bigoplus_{c} ((c \otimes \top) \amp A) \\
  & \cong (I \amp A) \oplus \bigoplus_{c} (c \otimes (c \lto A))
\end{align*}

Which effectively states that parses of a grammar $A$ are either empty or begin with some character from our fixed alphabet, a fact used in our proof of \cref{thm:dfa}.

\begin{theorem}
  \label{thm:dfa}
  For every regular grammar $g$, there is a DFA $D$ that is weakly equivalent to $g$.
\end{theorem}

\begin{proof}
  First, note that regular grammars have finitely many derivatives, up to weak equivalence \cite{?}. That is, we can isolate a finite set of words $w_{1},\dots,w_{n}$ such that $w \lto g$ is weakly equivalent to $w_{j} \to g$ for some $g$. These $w_{j}$ may then serve as representatives for their weak equivalences classes.

  The high-level idea underlying Brzozowski's DFA construction algorithm is that the states of the DFA are the derivative equivalence classes, and the transitions are formed by taking derivatives. If the derivative of $w \lto g$ with respect to a character $c$ is weakly equivalent to $w' \lto g$ then there is a transition from the state representing $w \lto g$ to the state representing $w' \lto g$. These pieces of data form the transition table form our DFA.\ Lastly, we make the state representing $w \lto g$ accepting if the type $I \amp (w \lto g)$ is inhabited. If it is inhabited the state for $w \lto g$ will have an $I$ as a summand, otherwise it will not. To represent this acceptance formally in our construction, use the function symbol $\mathsf{ac}(w) : 0 \oplus I$ to be $I$ if $w \lto g$ is inhabited and 0 otherwise. Similarly, $\mathsf{ac}(I)$ will represent the acceptance of the start state.

  We may precompute all this info and then construct the DFA by introducing states labeled $Y_{j}$ for each representative $w \in \{w_{1}, \dots, w_{n}\}$. Letting $j$ range in $\{s, 1, \dots, n\}$ and taking $w_{s} = I$ we can describe all states of the DFA with a similar structure.

  \[
  DFA(g) =
  \mu \begin{pmatrix}
        Y_{j} = \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes Y_{n_{j,c}} \\
  \end{pmatrix}.Y_{s}
\]

Note that by construction the target state of each transition, $Y_{n_{j,c}}$, is such that $c \lto (w_{j} \lto g)$ is weakly equivalent to $w_{n_{j, c}} \lto g$.

To turn a term of type $DFA(g)$ into a term of type $g$, we simply fold over the definitions of $DFA(g)$. Again, the structure of such a derivation,

\[
  \inferrule
  {
    \Gamma ; \Delta \vdash e : DFA(g) = \mu(Y_{j} = \hat{Y}_{j}).Y_{s} \\
    \Gamma ; x_{s} : \simulsubst {\mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes Y_{n_{s,c}}} {\gamma} = \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes B_{n_{s,c}} \vdash e_{s} : B_{s} \\
    \Gamma ;x_{j} : \simulsubst {\mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes Y_{n_{j,c}}}{\gamma} = \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes B_{n_{j,c}} \vdash e_{j} : B_{j}
  }
  {\Gamma ; \Delta \vdash \mathsf{mfold}(x_{s}.e_{s}, x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{s}}
\]

When instantiating the above rule, we make each $B_{k} = w_{k} \lto g$. In this manner, $B_{s} = I \lto g = g$ which is our desired output type. Let us examine one of the entailments to see how this is achievable. From a term $x_{j} : \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes (w_{n_{j,c}} \lto g)$, we need to produce some $e_{j} : w_{j} \lto g$. Which we may do in the following manner,

\begin{align*}
    \begin{aligned}
  e_{j} =
    \phi(\textbf{case}~x_{j}~\{&  \textbf{in}_{\mathsf{ac}} u \mapsto \mathbf{in}_{I \amp (w_{j} \to g)} u, \\
  & \textbf{in}_{c \otimes (w_{n_{j,c}} \lto g)} (v_{c} \otimes d_{j,c}) \mapsto \mathbf{in}_{c \otimes (w_{j} \lto g)}(v_{c} \otimes \alpha(d_{j,c}))
  \})
    \end{aligned}
\end{align*}

Where $\phi$ witnesses the isomorphism $g \cong (I \amp g) \oplus \bigoplus_{c} (c \otimes (c \lto g))$, and $\alpha$ is a function that acts a lookup table for replacing $d_{j,c} : w_{n_{j,c}} \lto g$ with an appropriate member of its weak equivalence class. In particular, when deriving $w_{j} \lto g$ we want $\alpha(d_{j,c}) : (w_{j} \otimes c) \lto g$. Finding such an element is indeed possible \emph{by construction of our DFA}. Recall that we added a transition between $Y_{j}$ and $Y_{n_{j,c}}$ exactly when $c \lto (w_{j} \lto g) = (w_{j} \otimes c) \lto g$ was weakly equivalent to $w_{n_{j,c}} \lto g$. That is, the multifold constructed above is well-defined, and will ultimately return a term of type $B_{s} = I \lto g = g$ so we do, in fact, have a derivation from $DFA(g)$ to $g$.

Now we wish to build a derivation from $g$ to $DFA(g)$.
\steven{fill in}
\end{proof}

@@

\begin{align*}
  r_{\oplus}(s_{\oplus}(e))
  & =_{\mu \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus \mu Z_{s}' \\
  &
    \begin{aligned}
      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
    \})))
    \end{aligned} \\
  &
    \begin{aligned}
    =_{\oplus \eta}
    \textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v)))
    \}))
    \end{aligned} \\
  &
    \begin{aligned}
    =
    \textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v)))
    \}
    \end{aligned} \\
  &
    \begin{aligned}
    =_{\mu \beta}
    \textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\textbf{in}_{g} \phi^{-1}(\mathsf{cons}_{\mu(X_{i}=\hat{X}_{i}).X_{s}}u')), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v'))
      \} \\
      & \text{~for }u' : \mu X_{s}', v' : \mu Z_{s}'
    \end{aligned} \\
  &
    \begin{aligned}
    =_{\oplus \beta}
    \textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha \phi \phi^{-1}(\mathsf{cons}(u')), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v'))
      \}
    \end{aligned} \\
  &
    \begin{aligned}
    =
    \textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha (\mathsf{cons}(u')), \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v'))
      \}
    \end{aligned} \\
   &
    \begin{aligned}
    =_{\mu \beta}
    \textbf{case}~e'~\{
    & \textbf{in}_{\mu X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{\mu X_{s}'} u) , \\
    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v)
      \}
    \end{aligned} \\
  & =_{\oplus \eta} \mathsf{cons}(e') \\
  & =_{\mu \beta} e
\end{align*}

@@

\[
  NFA(g \otimes g') = \mu
  \begin{pmatrix}
    Y_{s} = X_{s}' \\
    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
    X_{e}' = Z_{s}' \\
    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
    Z_{e}' = Y_{e} \\
    Y_{e} = I
  \end{pmatrix}.Y_{s}
\]
